The influence of habitat quality and environmental transmission on host-pathogen metapopulation dynamics

### 2014-11-02 Manuscript/manuscript.tex
Discussion edits
Edits to try to synthesize the insights from highvlow and the conditions
when high or low quality is better.

Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

These findings, that low quality habitat inhibits disease spread while high quality habitat enhances it, suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.  This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  Shifting the patch quality distribution towards high quality patches in our model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park, but by including habitat heterogeneity and metapopulation structure, we are able to investigate the role of individual patches and their contribution to total occupancy.

These contributions, and the relative strength of the above trade-offs, depend largely on the strength and structure of the different avenues of pathogen transmission.  Broadly, there are three different scenarios that emerge from this and the above studies: conditions that support very weak pathogen spread, conditions that facilitate intermediate/endemic pathogen spread, and conditions that produce easy/pandemic pathogen spread.  When pathogen spread is weak (e.g. due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.

However, in cases where the pathogen is able to spread more easily (i.e. at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).  In these cases, where pathogen longevity is intermediate, or dispersal is limited by metapopulation structure (e.g. a lattice), rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.  More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.  Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.

As pathogen longevity increases and environmental reservoirs become more widespread, the trade-off shifts and the addition of high quality patches is again a net benefit to overall occupancy.  In these situations, environmental reservoirs overwhelm the susceptible refuges on low quality patches, such that they remain contaminated even across infrequent colonization events.  Under these conditions of widespread infection, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, thereby increasing total occupancy relative to more balanced or low quality patch distributions.

@@

>>>>

Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Attracting colonists helps position high quality patches as the infectious "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for metapopulation occupancy vary depending on the structure and dynamics of the metapopulation.  Specifically, these findings, that low quality habitat inhibits disease spread while high quality habitat enhances it, suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.  This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  Shifting the patch quality distribution towards high quality patches in our model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park, but by including habitat heterogeneity and metapopulation structure, we are able to investigate the role of individual patches and their contribution to total occupancy.

These contributions, and the relative strength of the above trade-offs, depend largely on the role of connectivity, how it is affected by patch quality, and the resulting strength and structure of the different avenues of pathogen transmission.  Broadly, we see three different scenarios that roughly correspond to the three sections of the "u"-shaped relationships in Fig~\ref{highvlow}: conditions that support very weak pathogen spread, conditions that facilitate intermediate/endemic pathogen spread, and conditions that produce easy/pandemic pathogen spread.  When pathogen spread is weak (e.g. due to low longevity; the top left of the "u"), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.

However, in cases where the pathogen is able to spread more easily (i.e. at an endemic level; the trough of the "u"), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).  In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.  The shift in this trade-off can arise from conditions that change either side of that balance.  High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions, and thus for intermediate longevities, high quality distributions experience considerably greater pathogen spread and low quality habitat thus produce higher occupancy.

The trade-off can also switch in cases that disrupt the positive benefits of high quality habitat for metapopulation stability, i.e. with pathogens with high disease-induced mortality or where dispersal is limited by metapopulation structure.  In these cases, rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.  More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.  Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.

As pathogen longevity increases and environmental transmission becomes more important relative to direct transmission (i.e. the top right of the "u") the trade-off shifts back and the addition of high quality patches is again a net benefit to overall occupancy.  In these situations, environmental reservoirs overwhelm the susceptible refuges on low quality patches, such that they remain contaminated even across infrequent colonization events.  Under these conditions of widespread environmental transmission, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, with little additional pathogen spread, thereby increasing total occupancy relative to more balanced or low quality patch distributions.

@@

The different scenarios and parameterizations explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.  The consistent "u"-shaped trend in the relative benefit of low and high quality habitat as a function of pathogen longevity suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.  By evaluating the relative importance of direct versus environmental transmission, along with the relative effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g. plague in prairie dogs, or white nose syndrome in bats) fall on this spectrum and the corresponding consequences for management.
